---
title: "Report"
subtitle: "Exploring relationships between basic water usage, Hep-B vaccine availability, population and life expectancy"
author: "Harry Walsh"
date: 04/12/2024
format: html
toc: true
theme: minty
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)
library(plotly)
library(dplyr)

#data
unicef_indicator_1_2_ <- read_csv("C:/Users/Home/OneDrive/Desktop/Quarto/unicef_indicator_1 (2).csv")
unicef_indicator_2_2_ <- read_csv("C:/Users/Home/OneDrive/Desktop/Quarto/unicef_indicator_2 (2).csv")
unicef_metadata_2_ <- read_csv("C:/Users/Home/OneDrive/Desktop/Quarto/unicef_metadata (2).csv")
```

## Introduction
- Welcome to my Quarto Document. In this document, I will display a series of different **maps** and *charts*.
- For each **map** & *chart*, I will outline the information that is contained in the graphic.

## Excel File Number 1
The first excel file that I was given was titled ***unicef_indicator1***. The data being displayed was *"Proportion of population using basic drinking water services"*. This excel file had this data, as a %, for each country in the world, for each year. The time period ranged from **2000 to 2022**.

## Excel File Number 2
The second excel file that I was given was titled ***unicef_indicator2***. The data being displayed was *"Percentage of live births who received hepatitis-B-containing vaccine within 24 hours of birth"*. This excel file also had this data, as a %, for each country in the world, for each year. The time period ranged from **1960 to 2022**.

## Excel File Number 3
The third and final excel file that I was given was titled ***unicef_metadata***. This excel file was different to that of the above. There were a number of key metrics being displayed here including:
- Country
- Year (from 1960 - 2022)
- Population
- GNI (current US$)
- Life Expectancy


## World Map Chart
For my first visualisation, I designed a world map chart. I graphed the **Proportion of population using basic drinking water services**. From my graph, we can see which country has a high and low %, based off the colour their country is highlighted in.

```{r}
#| fig-cap: "World Map showing the Proportion of Population Using Basic Drinking Water Services - from 0% (dark blue) to 100% (light blue)"
map_world <- map_data("world")
map_data_join <- full_join(unicef_indicator_1_2_, map_world, by = c("country" = "region"))
ggplot(map_data_join) +
  aes(x = long, y = lat, group = group, fill = obs_value) +
  geom_polygon() +
  labs(
    main = "XYZ",
    x = "Longitude",
    y = "Latitude",
    fill = "% of 
    pop. 
    using 
    basic 
    water 
    services") +
  theme_classic()
```


## Time Series Chart
For my time series chart, I decided to focus on my home country - *Ireland*. I graphed the Proportion of population using basic drinking water services, in Ireland, from the year 2000, to 2022. It is evident from my graph, that this % took a ***sharp decline*** between the years 2005 to 2010. and has remained low ever since.


```{r}
#| fig-cap: "Proportion of Population Using Basic Drinking Water Services, in Ireland, from 2000 to 2022"
unicef_indicator_1_2_ %>%
  filter(country == "Ireland") %>%
  ggplot(aes(time_period, obs_value, color = country)) +
  geom_line() +
  theme_light() +
  labs(
    x = "Years",
    y = "% of pop. using basic water services"
  )
```



## Bar Chart
For my bar chart, I decided to filter my data. I graphed the ***Percentage of live births who received hepatitis-B-containing vaccine within 24 hours of birth***, in *Afghanistan, Albania, Algeria, Andorra and Argentina*. It is evident from my graph, that the % for each country differs quiet a lot - take a look for yourself!

```{r}
ggplot(unicef_indicator_2_2_) +
  aes(country2, value2, fill = "Red") +
  geom_col() +
  labs(
    x = "Country",
    y = "% of live births who received Hep-B vaccine",
    fill = "% of 
    Live 
    Births 
    that 
    Recieved 
    Hep-B 
    Vaccines") +
  theme_light()
```

## Scatter Plot
For my scatter-plot, I decided to again filter my data, choosing the countries *Afghanistan, Bhutan, Ireland, Albania, Peru, Spain*. I graphed the ***Life Expectancy vs Population***, of these chosen countries. From my graph, we can again see that the point for each country differs quiet a lot.

```{r}
#| fig-cap: "A scatter-plot mapping population and life expectancy of Afghanistan, Bhutan, Ireland, Albania, Peru and Spain. (Note: options(scipen = 999) was not used as the numbers wouldve layered on eachother)"
ggplot(unicef_metadata_2_) +
  aes(pop1, life.exp, fill = "Countries -
      Afghanistan,
      Bhutan,
      Ireland,
      Albania,
      Peru,
      Spain,") +
  geom_point() +
  labs(
    x = "Population",
    y = "Life Expectancy") +
  theme_dark()
```


## Conclusion
I feel as though it is our duty to fight these issues to our world leaders and raise more awareness.
I have you have enjoyed my document, and it gave you an insight into some of the issues our world faces. ***Thank You***.
